---
title: "Homework 1"
author: "Xinyi Shang (A69043536)"
date: today
format: 
  html:
    html-math-method: mathjax
editor: visual
---

## Package

```{r}
library(tidyverse)
library(bench)
library(palmerpenguins)
library(purrr)
library(ggplot2)
```

## Part1

### 1.1

### (a)

```{r}
x <- iris$Sepal.Length
vectorized_zscore <- function(x) {
  (x - mean(x)) / sd(x)
}

loop_zscore <- function(x) {
  z <- numeric(length(x))
  mu <- mean(x)
  s <- sd(x)
  
  for (i in seq_along(x)) {
    z[i] <- (x[i] - mu) / s
  }
  
  z
}

bm <- bench::mark(
  vectorized = vectorized_zscore(x),
  loop = loop_zscore(x),
  iterations = 50,
  check = TRUE
)

bm

```

### (b)

```{r}
bm_plot <- data.frame(
  method = as.character(bm$expression),
  median_time_ms = as.numeric(bm$median) * 1000
)
ggplot(bm_plot, aes(x = method, y = median_time_ms)) +
  geom_col() +
  labs(
    title = "Median Execution Time: Vectorized vs Loop Z-score",
    x = "Method",
    y = "Median execution time (ms)"
  )
```

### (c)

```{r}
head(ChickWeight)

ChickWeight_new <- ChickWeight %>%
  group_by(Chick) %>%
  mutate(
    Growth.Rate = (weight - lag(weight)) / (Time - lag(Time)),
    Weight.Gain.Percent = (last(weight) - first(weight)) / first(weight) * 100
  )

diet_summary <- ChickWeight_new %>%
  split(.$Diet) %>%
  map_dfr(~ tibble(
    Avg.Growth.Rate = mean(.x$Growth.Rate, na.rm = TRUE),
    Mean.Max.Weight = mean(tapply(.x$weight, .x$Chick, max),na.rm = TRUE)
  ), .id = "Diet")

diet_summary

```

## 1.2

### (a)

```{r}
library(palmerpenguins)
head(penguins)
species_list <- penguins %>%
  split(.$species) %>%
  map(~ {
    attr(.x, "sample_size") <- nrow(.x)
    .x
  })

species_list
attr(species_list$Adelie, "sample_size")
```

### (b)

```{r}
numeric_data <- penguins[, sapply(penguins, is.numeric)]

lapply(numeric_data, mean, na.rm = TRUE)

map_dbl(numeric_data, mean, na.rm = TRUE)

```

### （c)

```{r}
tapply(penguins$body_mass_g, penguins$species, mean, na.rm = TRUE)

tapply(penguins$body_mass_g, 
       list(penguins$species, penguins$sex), 
       mean, na.rm = TRUE)
```

### (d)

```{r}
x<-c(1,2,3)
y<-x
y[1]<-100
x
y
```

When `y <- x` is run, R does not immediately copy the data — both `x` and `y` point to the same vector in memory. Once `y` is modified, R creates a separate copy for `y` so that `x` remains unchanged. This is called copy-on-modify: R only copies data when it has to, which keeps results independent while saving memory.

### (e)

```{r}
system.time({
  result <- numeric(0)
for (i in 1:10000) {
  result <- c(result, i^2)
}
})

system.time({
  result <- numeric(10000)
for (i in 1:10000) {
  result[i]<- i^2
}
})
```

The preallocated version is faster because it creates the output vector only once and then fills in each position. The inefficient version repeatedly grows the vector with `c()`, which forces R to keep allocating new memory and copying old values. That repeated copying makes the loop much slower.

# part 2

## 2.1

```{r}
A <- matrix(c(4, 2, 1, 5), nrow = 2)
B <- matrix(c(3, 1, 0, 2), nrow = 2)
C <- matrix(c(1, 3, 5, 2, 4, 6), nrow = 2)
```

### (a)

```{r}
A + B
A * B
A %*% B
t(A) %*% C
solve(A)
```

1.  `A + B` adds the two matrices element by element.

2.  `A * B` multiplies the two matrices element by element.

3.  `A %*% B` performs matrix multiplication.

4.  `t(A) %*% C` first transposes `A` and then multiplies the result by `C`.

5.  `solve(A)` computes the inverse of `A` because `A` is square and invertible.

### （b)

```{r}
is_symmetric <- function(M) {
  if (nrow(M) != ncol(M)) {
    stop("Error: input matrix is not square")
  }
  isTRUE(all.equal(M, t(M)))
}

is_symmetric(A)              
is_symmetric(t(A) %*% A)     
is_symmetric(B)              
```

## 2.2

### (a)

```{r}
M <- matrix(c(4, 2, 2, 3), nrow = 2)
eig <- eigen(M)
eigenvalues <- eig$values
eigenvectors <- eig$vectors
print(eigenvalues)
print(eigenvectors)
```

### (b)

```{r}
for (i in 1:2) {
  x <- eigenvectors[, i]
  lambda <- eigenvalues[i]
  comparison <- cbind(M %*% x, lambda * x)
  colnames(comparison) <- c("M %*% x", "lambda * x")
  cat("--- Eigenvector", i, "---\n")
  print(comparison)
}

```

### （c)

```{r}
P <- eigenvectors
D <- diag(eigenvalues)
M_reconstructed <- P %*% D %*% solve(P)

print(M_reconstructed)
print(M)  
```

## 2.3

### (a)

```{r}
X <- matrix(c(3, 2, 2,
              2, 3, -2,
              2, -2, 3), nrow = 3, byrow = TRUE)

# SVD
svd_X <- svd(X)
svd_X

U <- svd_X$u
d <- svd_X$d
V <- svd_X$v

# Rank-1 approximation
X_rank1 <- d[1] * U[, 1] %*% t(V[, 1])

# Frobenius norm of the difference
fro_error <- norm(X - X_rank1, type = "F")

X_rank1
fro_error
```

The main information lost is that the matrix has a second equally large singular value, so a rank-1 approximation removes another major direction of variation instead of only removing a small residual component.

### （b)

```{r}
set.seed(228)
n <- 100
x <- runif(n, -5, 5)
y <- 3 + 2 * x + rnorm(n, sd = 1)

# Design matrix
X <- cbind(1, x)

# QR decomposition approach
qr_X <- qr(X)
Q <- qr.Q(qr_X)
R <- qr.R(qr_X)

beta_qr <- solve(R, t(Q) %*% y)
beta_qr

# Compare with lm()
model <- lm(y ~ x)
coef(model)

# Optional: put results side by side
comparison <- data.frame(
  Method = c("QR", "lm"),
  Intercept = c(beta_qr[1], coef(model)[1]),
  Slope = c(beta_qr[2], coef(model)[2])
)
comparison
```

QR decomposition provides a more numerically stable solution than normal equations $(X^TX)^{-1}X^Ty$ for least squares problems. While normal equations explicitly compute $X^TX$, which squares the condition number and amplifies numerical errors, QR maintains the original conditioning. This stability is especially crucial for ill-conditioned matrices commonly encountered in regression problems.

## Part 3

## 3.1

### (a)

```{r}
rosenbrock <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  (1 - x1)^2 + 100 * (x2 - x1^2)^2
}
```

### （b)

```{r}
rosenbrock_grad <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  
  c(
    -2 * (1 - x1) - 400 * x1 * (x2 - x1^2),
    200 * (x2 - x1^2)
  )
}
```

### (c)

```{r}
gradient_descent <- function(init, grad_fn, fn,
                             learning_rate = 0.001,
                             max_iter = 10000,
                             tol = 1e-6) {
  x <- init
  
  for (iter in 1:max_iter) {
    grad <- grad_fn(x)
    
    # stop if Euclidean norm of gradient is small enough
    if (sqrt(sum(grad^2)) < tol) {
      return(list(
        solution = x,
        value = fn(x),
        iterations = iter - 1,
        converged = TRUE
      ))
    }
    
    x <- x - learning_rate * grad
  }
  
  # if max_iter reached before convergence
  list(
    solution = x,
    value = fn(x),
    iterations = max_iter,
    converged = FALSE
  )
}
```

### (d)

```{r}
result <- gradient_descent(
  init = c(-1.2, 1.0),
  grad_fn = rosenbrock_grad,
  fn = rosenbrock,
  learning_rate = 0.001,
  max_iter = 10000,
  tol = 1e-6
)

result
```

The algorithm moved close to the true minimum (1,1), but with learning rate 0.001 it did not fully converge by 10000 iterations. The final function value was very small, so it was close to the optimum even though the stopping rule was not triggered.

## 3.2

### (a)

```{r}
# Starting point
x0 <- c(-1.2, 1.0)

# BFGS optimization
bfgs_result <- optim(
  par = x0,
  fn = rosenbrock,
  gr = rosenbrock_grad,
  method = "BFGS"
)

bfgs_result
```

Using `optim()` with method `"BFGS"`, the algorithm converged to `(1, 1)`, which is the true minimum of the Rosenbrock function. The final function value was essentially 0, so BFGS found the optimum very accurately.

### (b)

```{r}
# Nelder-Mead optimization
nm_result <- optim(
  par = x0,
  fn = rosenbrock,
  method = "Nelder-Mead"
)

nm_result

comparison_32 <- data.frame(
  Method = c("Gradient Descent", "BFGS", "Nelder-Mead"),
  x1 = c(result$solution[1], bfgs_result$par[1], nm_result$par[1]),
  x2 = c(result$solution[2], bfgs_result$par[2], nm_result$par[2]),
  FunctionValue = c(result$value, bfgs_result$value, nm_result$value),
  Iterations = unname(c(result$iterations, bfgs_result$counts[2], nm_result$counts[1])),
  FunctionEvaluations = unname(c(NA, bfgs_result$counts[1], nm_result$counts[1]))
)

comparison_32
```

Among the three methods, BFGS performed best overall. It converged to the true minimum `(1,1)` with a function value essentially equal to 0, so it had the highest solution accuracy. Nelder-Mead also found a solution very close to the optimum, but it required more function evaluations because it does not use gradient information. My gradient descent implementation moved close to the minimum, but it was slower and did not fully converge within the maximum number of iterations.

## 3.3

### （a)

```{r}
run_simulation <- function(n_samples, n_reps, dist_type) {
  # Initialize storage for this condition
  estimates <- matrix(
    NA,
    nrow = n_reps,
    ncol = 3,
    dimnames = list(NULL, c("mean", "median", "trimmed"))
  )
  
  # Run n_reps replications
  for (k in 1:n_reps) {
    # Generate data under the specified distribution
    if (dist_type == "normal") {
      x <- rnorm(n_samples, mean = 0, sd = 1)
    } else if (dist_type == "t3") {
      x <- rt(n_samples, df = 3)
    } else if (dist_type == "contaminated") {
      # 90% from N(0,1), 10% from N(0,10)
      n_clean <- rbinom(1, n_samples, 0.9)
      x_clean <- rnorm(n_clean, mean = 0, sd = 1)
      x_cont <- rnorm(n_samples - n_clean, mean = 0, sd = sqrt(10))
      x <- c(x_clean, x_cont)
    } else {
      stop("dist_type must be 'normal', 't3', or 'contaminated'")
    }
    
    # Compute estimators
    estimates[k, "mean"] <- mean(x)
    estimates[k, "median"] <- median(x)
    estimates[k, "trimmed"] <- mean(x, trim = 0.1)
  }
  
  # Return results as a data frame
  return(as.data.frame(estimates))
}
```

### (b)

```{r}
summarize_metrics <- function(estimates_df, true_value = 0, dist_name) {
  summary_df <- data.frame(
    distribution = character(),
    estimator = character(),
    bias = numeric(),
    variance = numeric(),
    mse = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (est in c("mean", "median", "trimmed")) {
    est_values <- estimates_df[[est]]
    
    bias_val <- mean(est_values) - true_value
    variance_val <- var(est_values)
    mse_val <- mean((est_values - true_value)^2)
    
    summary_df <- rbind(
      summary_df,
      data.frame(
        distribution = dist_name,
        estimator = est,
        bias = bias_val,
        variance = variance_val,
        mse = mse_val,
        stringsAsFactors = FALSE
      )
    )
  }
  
  return(summary_df)
}

set.seed(228)

# Run simulation for each distribution
sim_normal <- run_simulation(n_samples = 30, n_reps = 1000, dist_type = "normal")
sim_t3 <- run_simulation(n_samples = 30, n_reps = 1000, dist_type = "t3")
sim_cont <- run_simulation(n_samples = 30, n_reps = 1000, dist_type = "contaminated")

# Summarize performance
summary_normal <- summarize_metrics(sim_normal, true_value = 0, dist_name = "Normal(0,1)")
summary_t3 <- summarize_metrics(sim_t3, true_value = 0, dist_name = "t(df=3)")
summary_cont <- summarize_metrics(sim_cont, true_value = 0, dist_name = "Contaminated Normal")

# Combine results into one summary table
summary_table <- rbind(summary_normal, summary_t3, summary_cont)

# Round values for presentation
summary_table$bias <- round(summary_table$bias, 4)
summary_table$variance <- round(summary_table$variance, 4)
summary_table$mse <- round(summary_table$mse, 4)

summary_table

```

### 

(c) 

```{r}
summary_table[order(summary_table$distribution, summary_table$mse), ]
```

Under the normal distribution, the mean performed best because it had the smallest variance and MSE. Under the `t(df = 3)` distribution, the trimmed mean performed best, with the median close behind, while the mean performed worst because it was more sensitive to heavy tails. Under the contaminated normal distribution, the trimmed mean again had the smallest MSE, and both robust estimators outperformed the mean. Overall, the mean worked best for clean normal data, while the trimmed mean was the most reliable under heavy tails or contamination.

## 4.1

### （a)

```{r}
bootstrap_mean <- function(x, B = 2000, ci_type = c("percentile", "basic", "normal")) {
  ci_type <- match.arg(ci_type)
  
  n <- length(x)
  theta_hat <- mean(x)
  
  # Store bootstrap means
  boot_stats <- numeric(B)
  
  # Monte Carlo bootstrap
  for (b in 1:B) {
    x_star <- sample(x, size = n, replace = TRUE)
    boot_stats[b] <- mean(x_star)
  }
  
  # Bootstrap standard error
  se_boot <- sd(boot_stats)
  
  # 95% CI
  alpha <- 0.05
  z <- qnorm(1 - alpha / 2)
  q_lower <- quantile(boot_stats, probs = alpha / 2)
  q_upper <- quantile(boot_stats, probs = 1 - alpha / 2)
  
  if (ci_type == "percentile") {
    ci <- unname(c(q_lower, q_upper))
  } else if (ci_type == "basic") {
    ci <- unname(c(2 * theta_hat - q_upper, 2 * theta_hat - q_lower))
  } else if (ci_type == "normal") {
    ci <- c(theta_hat - z * se_boot, theta_hat + z * se_boot)
  }
  
  return(list(
    estimate = theta_hat,
    se_boot = se_boot,
    ci = ci,
    boot_stats = boot_stats,
    ci_type = ci_type
  ))
}

set.seed(42)
returns <- rnorm(50, mean = 0.08, sd = 0.12)
returns <- returns + rexp(50, rate = 20) - 0.05

set.seed(228)
boot_perc <- bootstrap_mean(returns, B = 2000, ci_type = "percentile")

set.seed(228)
boot_basic <- bootstrap_mean(returns, B = 2000, ci_type = "basic")

set.seed(228)
boot_normal <- bootstrap_mean(returns, B = 2000, ci_type = "normal")

boot_perc$estimate
boot_perc$se_boot

boot_perc$ci
boot_basic$ci
boot_normal$ci

ci_table <- data.frame(
  Method = c("Percentile", "Basic", "Normal"),
  Lower = c(boot_perc$ci[1], boot_basic$ci[1], boot_normal$ci[1]),
  Upper = c(boot_perc$ci[2], boot_basic$ci[2], boot_normal$ci[2])
)

ci_table
```

### (b)

```{r}
hist(
  boot_perc$boot_stats,
  breaks = 25,
  main = "Bootstrap Distribution of the Mean",
  xlab = "Bootstrap Means",
  col = "lightgray",
  border = "white"
)

# Original sample mean
abline(v = boot_perc$estimate, lwd = 2, col = "black")

# Percentile CI
abline(v = boot_perc$ci, col = "red", lwd = 2, lty = 2)

# Basic CI
abline(v = boot_basic$ci, col = "blue", lwd = 2, lty = 3)

# Normal CI
abline(v = boot_normal$ci, col = "darkgreen", lwd = 2, lty = 4)

legend(
  "topright",
  legend = c("Sample Mean", "Percentile CI", "Basic CI", "Normal CI"),
  col = c("black", "red", "blue", "darkgreen"),
  lwd = 2,
  lty = c(1, 2, 3, 4),
  bty = "n"
)

```

The returns data are somewhat right-skewed because they were generated by combining a normal component with an exponential component. For skewed data, the percentile interval is often the most appropriate bootstrap CI because it directly uses the empirical quantiles of the bootstrap distribution and naturally reflects asymmetry. The normal interval may be less appropriate because it forces symmetry around the sample mean. The basic interval is also bootstrap-based, but for these data the percentile method is the most intuitive and usually the most reasonable choice.

## 4.2

### (a)

```{r}
bootstrap_ttest <- function(x, y, B = 2000) {
  n1 <- length(x)
  n2 <- length(y)
  
  # Observed difference in means
  obs_diff <- mean(y) - mean(x)
  
  # Enforce null hypothesis by centering both groups to common mean
  pooled_mean <- mean(c(x, y))
  x_null <- x - mean(x) + pooled_mean
  y_null <- y - mean(y) + pooled_mean
  
  # Bootstrap null distribution
  null_diffs <- numeric(B)
  
  for (b in 1:B) {
    x_star <- sample(x_null, size = n1, replace = TRUE)
    y_star <- sample(y_null, size = n2, replace = TRUE)
    null_diffs[b] <- mean(y_star) - mean(x_star)
  }
  
  # Two-sided bootstrap p-value
  p_boot <- mean(abs(null_diffs) >= abs(obs_diff))
  
  return(list(
    obs_diff = obs_diff,
    null_diffs = null_diffs,
    p_value = p_boot
  ))
}

set.seed(123)
control <- rnorm(40, mean = 10, sd = 2)
treatment <- rnorm(40, mean = 11.5, sd = 2)

set.seed(228)
boot_test <- bootstrap_ttest(control, treatment, B = 2000)

boot_test$obs_diff
boot_test$p_value
```

```{r}
hist(
  boot_test$null_diffs,
  breaks = 30,
  main = "Bootstrap Null Distribution of Difference in Means",
  xlab = "Difference in Means (Treatment - Control)",
  col = "lightblue",
  border = "white"
)

abline(v = boot_test$obs_diff, col = "red", lwd = 2)
abline(v = -abs(boot_test$obs_diff), col = "darkred", lwd = 2, lty = 2)
abline(v = abs(boot_test$obs_diff), col = "darkred", lwd = 2, lty = 2)

legend(
  "topright",
  legend = c("Observed Difference", "Two-sided cutoff"),
  col = c("red", "darkred"),
  lwd = 2,
  lty = c(1, 2),
  bty = "n"
)
```

### (b)

```{r}
t_test_result <- t.test(treatment, control, var.equal = FALSE)
t_test_result
```

The bootstrap test and the standard t-test should give similar results here because the data were generated from roughly normal distributions and the sample sizes are the same. In this case, the t-test is simpler and works well. The bootstrap approach is still helpful because it relies less on distribution assumptions. If the data were more skewed or had outliers, the bootstrap method would probably be more useful.
